---
title: "Bárðarbunga activities in a nutshell"
author: "Einar Hjörleifsson"
date: "17/09/2014"
output:
  html_document:
    toc: yes
---

```{r setup, echo=FALSE}
opts_chunk$set(message=FALSE,
               tidy=TRUE,
               tidy.opts=list(width.cutoff=55),
               pretty=FALSE,
               echo=FALSE,
               eval=TRUE,
               cache=FALSE,
               cache.path='cache/')
```

# The required packages

```{r}
require(ggplot2)
require(ggmap)
require(rgdal)
require(raster)
```
# Getting the data

## The background map

The background map is generate from raster objects (see bottom of document):

```{r}
eldar <- raster("map/eldar.img")
ext_eldar <- extent(eldar)
X.MIN <- ext_eldar@xmin / 1e3
X.MAX <- ext_eldar@xmax / 1e3
Y.MIN <- ext_eldar@ymin / 1e3
Y.MAX <- ext_eldar@ymax / 1e3


eldar <- rasterToPoints(eldar)
eldar <- data.frame(eldar)
colnames(eldar) <- c("x","y","hill")
eldar$x <- eldar$x/1e3 - X.MIN
eldar$y <- eldar$y/1e3 - Y.MIN

water <- raster("map/water.img")
water <- rasterToPoints(water)
water <- data.frame(water)
water$x <- water$x/1e3 - X.MIN
water$y <- water$y/1e3 - Y.MIN

symbol <- data.frame(x=rep(40,5),y=rep(5,5),size=1:5)

attach("shapes.RData")

theme_nothing2 <- function() {
  return(theme(line = element_blank(),
               rect = element_blank(), 
               text = element_blank(),
               axis.ticks.length = unit(0,"cm"),
               axis.ticks.margin = unit(0.01, "cm"), 
               legend.position = "none",
               panel.margin = unit(c(0,0,0,0), "cm"),
               plot.margin = unit(c(-1,-1,-1,-1), "cm"),
               complete = TRUE))
  }
eldar_base_plot <- ggplot() +
  theme_nothing2() +
  geom_raster(data=eldar,aes(x,y,fill=hill)) +
  geom_raster(data=water,aes(x,y),fill="white",alpha=0.25) +
  scale_fill_gradientn(colours=grey(1:100/100),guide="none") +
  geom_polygon(data=gangur,aes(x,y),fill="white",alpha=0.05) +
  geom_polygon(data=bunga,aes(x,y),fill="white",alpha=0.05) +
  coord_equal() +
  annotate("text",x=2,y=44,hjust=0,label="Base map: www.lmi.is",size=5) +
  annotate("text",x=2,y=42.5,hjust=0,label="DATA source: www.vedur.is",size=5) +
  annotate("text",x=2,y=40,hjust=0,label="fishvise",colour="white",size=7) +
  annotate("text",x=2,y=35,hjust=0,label="Recent activity",colour="red",size=7) +
  annotate("text",x=2,y=33.5,hjust=0,label="Historical trail",colour="blue",size=7) +
  geom_point(data=symbol[symbol$size %in% c(1,3,5),],aes(x,y,size=size*6),shape=1,col="red") +
  #scale_shape(solid=FALSE) +
  #geom_text(data=symbol,aes(x,y+(size-1)/1.75,label=size),size=3) +
  geom_text(data=symbol[symbol$size %in% c(1,3,5),],aes(x,y+(size-1)/1.35,label=size),size=7) +
  annotate("segment",x=25,xend=35,y=1.5,yend=1.5) +
  annotate("text",x=30,y=2,label="10 km") +
  scale_size_identity()


eldar_base_plot
```

## The earthquakes
Checked earthquake data reside the [Icelandic Meterological website](http://hraun.vedur.is/ja/viku/). 

First we define a little function, which allows us to read the year we may be interested in as well as the weeks. The function will also convert the long and lat data to kilometers.

```{r, eval=FALSE}
get_hraun <- function(weeks,year) {
  i <- weeks < 10
  weeks[i] <- paste("0",weeks[i],sep="")
  for (i in 1:length(weeks)) {
    tmp <- read.table(paste("http://hraun.vedur.is/ja/viku/",year,"/vika_",weeks[i],"/listi",sep=""),
                      header=TRUE,colClasses=c("integer",rep("character",2),rep("numeric",5)))
    if(i==1) {
      x <- tmp
    } else {
      x <- rbind(tmp,x)
    }
  }
  names(x) <- c("id","date","time","lat","lon","z","size","ML")
  x$date <- ymd_hms(paste(x$date,x$time,"UTC"))
  
  require(rgdal)
  #require(rgisland)
  PRO <- CRS('+proj=longlat +datum=WGS84')
  ISN94 <- CRS('+proj=lcc +lat_1=64.25 +lat_2=65.75 +lat_0=65 +lon_0=-19 +x_0=500000 +y_0=500000 +ellps=GRS80 +units=m +no_defs')
  
  x$id <- 1:nrow(x)
  x.stdf <- SpatialPointsDataFrame(x[,c('lon','lat')],
                                 x[,c('id','date')],
                                 proj4string=PRO)
  x.stdf <- spTransform(x.stdf,ISN94)
  x.df <- as.data.frame(x.stdf)
  names(x.df) <- c("id","date","x","y")
  x <- join(x,x.df[,c("id","x","y")])
  
  return(x)
}


get_hraun_recent <- function() {
  library(XML)
  library(RCurl)
  require(httr)
  url <- "http://hraun.vedur.is/ja/skjalftar/skjlisti.html"
  x2 <- GET(url)
  x2 <- readHTMLTable(rawToChar(x2$content), stringsAsFactors = F)[[3]]
  names(x2) <- c("date","time","lat","lon","z","size","quality","loc1","loc2","loc3")
  require(lubridate)
  require(stringr)
  x2$date <- ymd_hms(paste(ymd(x2$date),str_sub(x2$time,1,8),"UTC"))
  x2$lat <- as.numeric(str_replace(x2$lat,",","."))
  x2$lon <- as.numeric(str_replace(x2$lon,",","."))
  x2$z <- as.numeric(str_replace(x2$z,",","."))
  x2$size <- as.numeric(str_replace(x2$size,",","."))
  x2$quality <- as.numeric(str_replace(x2$quality,",","."))
  x2$st <- paste(x2$loc,x2$loc2,x2$loc3)
  x2 <- x2[,c("date","lon","lat","z","size","quality","st")]
  x <- x2
  # convert lon and lat to meters
  require(rgdal)
  PRO <- CRS('+proj=longlat +datum=WGS84')
  ISN94 <- CRS('+proj=lcc +lat_1=64.25 +lat_2=65.75 +lat_0=65 +lon_0=-19 +x_0=500000 +y_0=500000 +ellps=GRS80 +units=m +no_defs')
  
  x$id <- 1:nrow(x)
  x.stdf <- SpatialPointsDataFrame(x[,c('lon','lat')],
                                   x[,c('id','date','z','size','quality','st')],
                                   proj4string=PRO)
  
  x.stdf <- spTransform(x.stdf,ISN94)
  
  x.df <- as.data.frame(x.stdf)
  names(x.df) <- c("id","date","z","size","quality","st","x","y")
  x <- join(x,x.df[,c("id","x","y")])
  }
```

```{r, eval=FALSE}
haun <- get_hraun(weeks=33:37,year=2014)
#save(hraun,file="hraun.RData")
```

```{r}
attach("hraun.RData")
hraun_recent <- get_hraun_recent()
#i <- hraun_recent$date > max(hraun$date) & hraun_recent$quality >= 65
#cn <- c("date","x","y","z","size")
#hraun <- rbind(hraun[,cn],hraun_recent[i,cn])

hraun$x <- hraun$x / 1e3 - X.MIN
hraun$y <- hraun$y / 1e3 - Y.MIN

i <- hraun$x >= 0 & hraun$x <= 45 &
  hraun$y >= 0 & hraun$y <= 45 &
  year(hraun$date) == 2014 & month(hraun$date) %in% c(8,9)
d <- hraun[i,]
d <- d[month(d$date) %in% 8 & day(d$date) >= 14 | month(d$date) == 9,]
d <- d[d$size > 0,]
minute(d$date) <- second(d$date) <- 0

# define bunga and gangur
d$area <- NA
require(geo)
i <- geoinside(d,gangur,col.names = c("y","x"),option = 3)
d$area[i==1] <- "Dyke"
i <- geoinside(d,bunga,col.names=c("y","x"),option=3)
d$area[i==1] <- "Bárðarbunga"
Time <- min(d$date)
for (i in 1:1000) Time <- c(Time,Time[i]+hours(1))
Time <- Time[Time <= max(d$date)]
file_name <- str_sub(as.character(Time),1,16)
```


```{r}
plot_eldar <- function(d,i,Time) {
  # Add historical trail, all earthquakes up to now
  p <- eldar_base_plot + 
    geom_point(data=d[d$date <= Time[i],],aes(x,y),size=2,col="blue",alpha=0.3)
  
  # Fade out the last 10 hours
  x <- d[d$date %in% c(Time[max(1,i-9)]:Time[i]),]
  x <- x[order(x$date,decreasing=TRUE),]  
  Dates <- data.frame(date=unique(x$date))
  Dates$fade <- rev(as.numeric(max(Dates$date) - Dates$date) / (60*60))/10 + 0.1
  x <- join(x,Dates,by="date")
  
  # only plot size >= 1 for the recent activity
  j <- x$size >= 0
  if(any(j))  {
      p <- p + 
        geom_point(data=x[j,],aes(x,y,alpha=fade,size=size*6),shape=1,col="red") 
        #scale_size_identity() +
        #scale_shape(solid = FALSE)
      p <- p + geom_point(data=x[j,],aes(x,y,alpha=fade,size=size),col="red")
  }

  p <- p + annotate("text",x=2,y=37,hjust=0,label=file_name[i],size=10,col="white")
  return(p)
  }


plot_depth_x <- function(d,i,Time) {
  p <- ggplot() +
    theme_nothing2() +
    coord_equal()
  d$z <- ifelse(d$z > 20,20,d$z)
  p <- p + geom_point(data=d[d$date <= Time[i],],aes(x,-z),size=2,col="blue",alpha=0.3)
  # Fade out the last 10 hours
  x <- d[d$date %in% c(Time[max(1,i-9)]:Time[i]),]
  x <- x[order(x$date,decreasing=TRUE),]
  Dates <- data.frame(date=unique(x$date))
  Dates$fade <- rev(as.numeric(max(Dates$date) - Dates$date) / (60*60))/10 + 0.1
  x <- join(x,Dates,by="date")
  # only plot size >= 1 for the recent activity
  j <- x$size >= 0
  if(any(j))  {
    p <- p +
      geom_point(data=x[j,],aes(x,-z,alpha=fade,size=size*6),shape=1,col="red")
    p <- p + geom_point(data=x[j,],aes(x,-z,alpha=fade,size=size),col="red")
    }
  p <- p + scale_size_identity() + labs(x="",y="") + theme(legend.position="none")
  p <- p + scale_x_continuous(limits=c(0 ,45)) +
    scale_y_continuous(limits=c(-20,0))
  p <- p + annotate("segment", x = 0, xend = 0, y = 0, yend = -20, colour = "black")
  p <- p + annotate("text",x=0.5,y=0,hjust=0,label="0 km",size=5)
  p <- p + annotate("text",x=0.5,y=-5,hjust=0,label="5 km",size=5)
  p <- p + annotate("text",x=0.5,y=-10,hjust=0,label="10 km",size=5)
  p <- p + annotate("text",x=0.5,y=-15,hjust=0,label="15 km",size=5)
  p <- p + annotate("text",x=0.5,y=-20,hjust=0,label="20 km",size=5) #+
   #theme(panel.margin = unit(0, "cm"))
  return(p)
  }
plot_depth_x(d,300,Time) + coord_equal()

plot_depth_y <- function(d,i,Time) {
  p <- ggplot() +
    theme_nothing2() +
    coord_equal()
  d$z <- ifelse(d$z > 20,20,d$z)  
  p <- p + geom_point(data=d[d$date <= Time[i],],aes(z,y),size=2,col="blue",alpha=0.3)
  
  # Fade out the last 10 hours
  x <- d[d$date %in% c(Time[max(1,i-9)]:Time[i]),]
  x <- x[order(x$date,decreasing=TRUE),]  
  Dates <- data.frame(date=unique(x$date))
  Dates$fade <- rev(as.numeric(max(Dates$date) - Dates$date) / (60*60))/10 + 0.1
  x <- join(x,Dates,by="date")
  
  # only plot size >= 1 for the recent activity
  j <- x$size >= 0
  if(any(j))  {
      p <- p + 
        geom_point(data=x[j,],aes(z,y,alpha=fade,size=size*6),shape=1,col="red") 
      p <- p + geom_point(data=x[j,],aes(z,y,alpha=fade,size=size),col="red")
  }
  p <- p + scale_size_identity() #+ labs(x="",y="") + theme(legend.position="none")
  p <- p + scale_x_continuous(limits=c(0,20)) +
    scale_y_continuous(limits=c(0,45)) 
  p <- p + annotate("segment", x = 0, xend = 20, y = 0, yend = 0, colour = "black")
  p <- p + annotate("text",x=0,y=0.5,hjust=0,label="0 km",size=5,angle=90)
  p <- p + annotate("text",x=5,y=0.5,hjust=0,label="5 km",size=5,angle=90)
  p <- p + annotate("text",x=10,y=0.5,hjust=0,label="10 km",size=5,angle=90)
  p <- p + annotate("text",x=15,y=0.5,hjust=0,label="15 km",size=5,angle=90)
  p <- p + annotate("text",x=20,y=0.5,hjust=0,label="20 km",size=5,angle=90) 
  return(p)
  }

plot_activity <- function(d,i,Time) {
  Days <- Time
  hour(Days) <- 0
  Days <- unique(Days)
  
  dummy <- data.frame(date=rep(max(Days),2),n=c(271,75),area=c("Dyke","Bárðarbunga"))
  
  dummy_label <- data.frame(date=rep(max(Days)),n=c(265,70),area=c("Dyke","Bárðarbunga"))

  x <- d[d$date <= Time[i] & d$area %in% c("Dyke","Bárðarbunga"),]
  hour(x$date) <- 0

  x0 <- ddply(x,c("date","area"),summarise,n=length(date))
  x3 <- ddply(x[x$size >= 3,],c("date","area"),summarise,n=length(date))
  p <- ggplot() + 
    scale_x_datetime(limits=c(min(Days),max(Days))) +
    geom_linerange(data=dummy,aes(date,ymin=0,ymax=n),col="white",alpha=0.01)  +
    geom_linerange(data=x0,aes(date,ymin=0,ymax=n),col="grey",lwd=3)  +
    #theme_bw() +
    facet_grid(area ~ .,scale="free_y") +
    geom_text(data=dummy_label,aes(date,n,label=area),hjust=1,size=7) +
    theme(line = element_blank(),
          rect = element_blank(),
          #text = element_blank(),
          #axis.ticks.length = unit(0, "cm"),
          #axis.ticks.margin = unit(0.01, "cm"), 
          legend.position = "none",
          #panel.margin = unit(0, "lines"),
          plot.margin = unit(c(0, 0, -0.5, -0.5), "lines"),
          axis.text = element_text(size = 12),
          axis.text = element_text(size = 12),
          complete = TRUE) +
    labs(x="",y="",title="Quakes per day. Red: quakes >=3")
  if(nrow(x3) !=0) p <- p + geom_linerange(data=x3,aes(date,ymin=0,ymax=n),col="red",lwd=3) 
  return(p)
  }
```


## The view

```{r, eval=FALSE}
library(grid)
vplayout <- function(x, y) viewport(layout.pos.row = x, layout.pos.col = y)

i <- 700
jpeg(filename="test.jpg",width=950,height=950,bg="white")
grid.newpage()
pushViewport(viewport(layout = grid.layout(nrow=2,ncol=2,
                                           widths=c(0.6923077,1-0.6923077),
                                           heights=c(0.6923077,1-0.6923077),
                                           just=c("left","bottom"))))
print(plot_eldar(d,i,Time), vp = vplayout(1, 1))
print(plot_depth_x(d,i,Time), vp = vplayout(2,1))
print(plot_depth_y(d,i,Time), vp = vplayout(1,2))
print(plot_activity(d,i,Time),vp = vplayout(2,2))
dev.off()




for (i in 1:length(Time)) {
  jpeg(filename=paste("fig/",file_name[i],".jpg",sep=""),width=950,height=950,bg="white")
  grid.newpage()
  pushViewport(viewport(layout = grid.layout(nrow=2,ncol=2,
                                             widths=c(0.6923077,1-0.6923077),
                                             heights=c(0.6923077,1-0.6923077),
                                             just=c("left","bottom"))))
  print(plot_eldar(d,i,Time), vp = vplayout(1, 1))
  print(plot_depth_x(d,i,Time), vp = vplayout(2,1))
  print(plot_depth_y(d,i,Time), vp = vplayout(1,2))
  print(plot_activity(d,i,Time),vp = vplayout(2,2))
  dev.off()
  }
  


```

```{r, eval=FALSE}
system("convert -delay 15 *.jpg bunga_august.mp4")
```



# The background rasters

```{r, eval=FALSE}
eldar <- raster("/home/einarhj/stasi/gis/LMI/HILLSHADE_LMI_ISN93/hillshade_LMI_isn93.img")
ext_eldar <- extent(562969.5,613752.8,452149.4,493070.2)
eldar <- crop(eldar,ext_eldar)
eldar <- aggregate(eldar,4,mean)
writeRaster(eldar,file="background_map/eldar.img")

dsn <- "/home/einarhj/stasi/gis/LMI/IS50V_VATNAFAR_01072013_ISN93/IS50V_VATNAFAR_SHP"
water <- readOGR(dsn,"is50v_vatnafar_flakar_01072013")
r <- raster(nrow=nrow(eldar),ncol=ncol(eldar))
extent(r) <- ext_eldar
water <- rasterize(water, r, 1, fun="first")
writeRaster(water,file="background_map/water.img",overwrite=TRUE)
```


# area defintion
```{r}
GANGUR <- data.frame(lon=c(-17.45262,-17.26731,-16.91901,-16.96143,-16.82971,-16.74040),
                     lat=c( 64.64699, 64.58418, 64.72902, 64.74184, 64.88540, 65.02768))
GANGUR <- GANGUR[2:5,]
PRO <- CRS('+proj=longlat +datum=WGS84')
ISN94 <- CRS('+proj=lcc +lat_1=64.25 +lat_2=65.75 +lat_0=65 +lon_0=-19 +x_0=500000 +y_0=500000 +ellps=GRS80 +units=m +no_defs')

GANGUR$id <- 1:nrow(GANGUR)
GANGUR$name <- c("Bunga","g1_start","g1_end","g2_start","g2_end","askja")

# trial
#   make points along the line
GANGUR.l <- Line(GANGUR[,c('lon','lat')])
GANGUR.sl <- SpatialLines(list(Lines(GANGUR.l,ID='gangur')),proj4string=PRO)
GANGUR.pt <- spsample(GANGUR.sl,n=1000,type='regular')
GANGUR.df <- as.data.frame(coordinates(GANGUR.pt))
GANGUR.df$id <- GANGUR.df$name <- 1:nrow(GANGUR.df)
GANGUR.stdf <- SpatialPointsDataFrame(GANGUR.df[,c('lon','lat')],
                                 GANGUR.df[,c('id','name')],
                                 proj4string=PRO)
GANGUR.stdf <- spTransform(GANGUR.stdf,ISN94)
require(rgeos)
GANGUR.stdf <- gBuffer(GANGUR.stdf,width =  2500,quadsegs=100)
gangur <- fortify(GANGUR.stdf)
gangur <- gangur[gangur$group %in% "buffer.1", c("long","lat")]
names(gangur) <- c("x","y")
gangur$x <- gangur$x/1e3 - x.min
gangur$y <- gangur$y/1e3 - y.min
eldar_base_plot + geom_path(data=gangur,aes(x,y))

BUNGA <- data.frame(x=c(571344.5,574470.3),y=c(458781.6,461355.7))
BUNGA.l <- Line(BUNGA[,c('x','y')])
BUNGA.sl <- SpatialLines(list(Lines(BUNGA.l,ID='bunga')),proj4string=ISN94)
BUNGA.pt <- spsample(BUNGA.sl,n=1000,type='regular')
BUNGA.df <- as.data.frame(coordinates(BUNGA.pt))
BUNGA.df$id <- BUNGA.df$name <- 1:nrow(BUNGA.df)
BUNGA.stdf <- SpatialPointsDataFrame(BUNGA.df[,c('x','y')],
                                 BUNGA.df[,c('id','name')],
                                 proj4string=ISN94)
BUNGA.stdf <- gBuffer(BUNGA.stdf,width =  7000,quadsegs=100)
bunga <- fortify(BUNGA.stdf)
bunga <- bunga[, c("long","lat")]
names(bunga) <- c("x","y")
bunga$x <- bunga$x/1e3 - x.min
bunga$y <- bunga$y/1e3 - y.min
eldar_base_plot + geom_path(data=gangur,aes(x,y)) + geom_path(data=bunga,aes(x,y))

save(gangur,bunga,file="shapes.RData")
```
